---
title: "Reproducible Research - Peer Reviewed Assignment 01"
author: "LALIT KUMAR JIWANI"
date: "August 5, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```


```{r loading library, message=FALSE, warning=FALSE}

library(dplyr)
library(ggplot2)
library(Hmisc)

```

## Loading and preprocessing the data ##
####################################################

```{r Loading and Preprocessing the dat}

if(!file.exists('activity.csv')){
  unzip('repdata_data_activity.zip')
}

data_pac <- read.csv("activity.csv") ##data about Personal movement using ACtivity 

```

# PART 01 #
## What is mean total number of steps taken per day? ##

```{r }

dates <- group_by(data_pac, date)

data_sum_step <- summarise(dates, steps_in_day = sum(steps, na.rm = TRUE))
plot(data_sum_step$date, data_sum_step$steps_in_day, type = "l", xlab="DATE", ylab="Number of steps", main = "Sum of steps in Day")


```

### Make a histogram of the total number of steps taken each day ###
```{r }
hist(data_sum_step$steps_in_day, breaks = 50, xlab = "Number of steps in a Day", ylab = "Frequency", main = "Total Number of Steps") 
#number of breaks is chosen for better distinguishability 
```


### Calculate and report the mean and median of the total number of steps taken per day
```{r }
mean_total_no_steps <- mean(data_sum_step$steps_in_day)
median_total_no_steps <- median(data_sum_step$steps_in_day)
```

Mean number of steps = `r mean_total_no_steps`

Median number of steps = `r median_total_no_steps`


# PART 02 #################################
## What is the average daily activity pattern?

### Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r}
data_intervals <- group_by(data_pac, interval)

data_mean_interval <- summarise(data_intervals, steps_in_interval = mean(steps, na.rm = TRUE))
plot(data_mean_interval$interval, data_mean_interval$steps_in_interval, type = "l", xlab = "Time Interval", ylab = "Mean Number of Steps", main = "Mean Steps in Interval")

```


### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r }
max_steps_interval_index <- which.max(data_mean_interval$steps_in_interval)
max_steps_interval_value <- data_mean_interval$interval[max_steps_interval_index]
```

 5-minute interval containing max number steps =  `r max_steps_interval_value`
 
# PART 03 ##

###	Imputing the missing valus 

```{r}
new_DF <- data_pac[rowSums(is.na(data_pac)) > 0,]
number_of_rows_without_NA <- nrow(new_DF)
```
Total number of missing values in the dataset = `r number_of_rows_without_NA`

###	Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
impute_data_pac <- data_pac

impute_data_pac$steps <- impute(data_pac$steps, fun=mean)
```
impute_data_pac is the new data

###	Make a histogram of the total number of steps taken each day

```{r}
impute_data_pac_dates <- group_by(impute_data_pac, date)

##
impute_data_sum_step <- summarise(impute_data_pac_dates, steps_in_day = sum(steps, na.rm = TRUE))
hist(impute_data_sum_step$steps_in_day, breaks = 50, xlab = "Number of steps in a Day", ylab = "Frequency", main = "Total Number of Steps in Imputed data")
```


```{r}
mean_total_no_steps_imputed <- mean(impute_data_sum_step$steps_in_day)
median_total_no_steps_imputed <- median(impute_data_sum_step$steps_in_day)
```
Mean number of steps in imputed data = `r mean_total_no_steps_imputed`

Median number of steps in imputed data = `r median_total_no_steps_imputed`

As can be seen that these values differ from the original data.

# PART 04 ##

##Are there differences in activity patterns between weekdays and weekends?

###Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
pdate <-  as.POSIXlt(impute_data_pac$date)
impute_data_pac$dayType <-  ifelse(pdate$wday %in% c(0,6), "weekday", "weekend")
```

### Make a panel plot containing a time series plot of the 5-minute interval

```{r}
impute_data_pac_mean <- aggregate(steps ~ interval + dayType, data = impute_data_pac, mean)
ggplot(impute_data_pac_mean, aes(interval, steps)) +
  geom_line() +
  facet_grid(dayType ~ .) +
  xlab("5-minute interval") +  
  ylab("avarage number of steps")
```

